
{\noindent \huge \bf Acknowledgments\\}

I want to extend my thanks to my advisors Frans Kaashoek and Adam Belay
for their guidance and mentorship, to my other committee members Nickolai Zeldovich and Mengjia Yan for their valuable feedback, and to all the past and present members of PDOS who I have learned so much from over these years.
Thank you.

I would also like to thank all of my collaborators including Jack Cook, Jules Drean, Anton Cao, Cel Skeggs, Amy Ousterhout, Joshua Fried, Hari Balakrishnan, Jon Gjengset, Malte Schwarzkopf, Lara Timbó Araújo, Martin Ek, Eddie Kohler, Robert Morris, Sagar Jha, Ken Birman, and Edward Tremel. Your insights and support have meant a ton.

I am extraordinarily thankful to my friends and family have supported me every step of this journey.
You have shaped my experiences, given me meaning and purpose, and helped me drive positive change.
I would not be where I am today without you.

\begin{center}
    ~\\[0.5\baselineskip]
    $\star~\star~\star$
    ~\\[\baselineskip]
\end{center}


\noindent
This dissertation includes work from the following papers:
\begin{itemize}
\item Efficiently mitigating transient execution attacks using the unmapped speculation contract. 
    Jonathan Behrens, Anton Cao, Cel Skeggs, Adam Belay, M. Frans Kaashoek, and Nickolai Zeldovich.
    \textit{OSDI 2020}.
\item Performance Evolution of Mitigating Transient Execution Attacks. 
    Jonathan Behrens, M. Frans Kaashoek, and Adam Belay.
    To appear. \textit{\mbox{EuroSys} 2022}.
\end{itemize}
